---
title: "Play Videos in R"
format: 
  html:
    toc: true
    toc-depth: 4
---

This bite has some quick code to parse video files and render them to the console using only vanilla methods. (and the `av` package to parse video)

Code for this can be found in `video-player.R` in the [repo](https://github.com/gbkorr/r-bites/blob/main/video-player.R) or in the [main r-bites zip](https://github.com/gbkorr/r-bites/releases).

```{r eval=F}
vid = load.video('XcQ.mp4')
play.video(vid)
```

The console **must be resized** (and zoomed in or out) until the video renders properly.

::: {.callout-note title="Demo" collapse="true"}
{{< video video-player-demo.mov >}} Squinting helps a lot.
:::


## Big Picture

To render a video, we need:

-   a way to convert high-res, color images to something viewable in the console

-   a way to load and display each frame at the right time

## Functions

### **render.matrix()**

Renders an integer matrix to the console as a single wrapped line, overwriting the previous render.matrix() output. This allows us to animate video frames by drawing them in succession.

### **render.image()**

Converts an image (e.g. from `jpeg::readJPEG()`) into a matrix suitable for `render.matrix`. This grayscales and quantizes it to only a few shades of gray.

### **load_video()**

Reads a video file and generates an image for each frame using `av::av_video_images()`. The video is also downscaled and set to a specified framerate for ease of rendering.

### **play_video()**

Reads a list of images as generated by `load.video()` or `av::av_video_images()` and renders each frame at the appropriate time. If rendering takes too long, frames will be skipped to preserve the overall video speed.

## Code Details

### **Libraries**

```{r eval=F}
library(av)
library(jpeg)
```

We use the standard video processing package `av` to generate JPEG images from a video file, and the `jpeg` package to read these images.

`av` uses [FFmpeg](https://www.ffmpeg.org/download.html) to process videos, which you might have to install.

___

### **render.matrix()**

::: {.callout-tip title="code" collapse="true"}
```{r eval=F}
render.matrix = function(M,palette=c('▓',' ')) {
	tab_buffer = matrix(-1,nrow(M),10)
	#matrix of integers to indicate tab chars
	#these go to the right of the matrix
	
	period_buffer = -2
	#a column of periods is also added on the left
	#otherwise, leading whitespace gets removed
	
	new_matrix = cbind(period_buffer,M,tab_buffer)
	new_matrix = t(new_matrix) + 3
	#transpose to orient correctly
	#add 3 to make it 1-index
	
	cat(
		#clear previous render
		'\r',		
		#index palette to convert matrix to characters
		c('.','\t',palette)[new_matrix], sep='')
}
```
:::

Console rendering is achieved by calling `base::cat()` on a single-line string which wraps to match the dimensions of the matrix we feed in.

Additionally, we:

-   use `\r` to clear the previous frame rendered by the function

-   use `\t` tab characters to wrap each row onto a new line, *without* having to use actual `\n` newlines (which would break `\r`)

The input matrix consists of positive integers (and 0) corresponding to the index in `palette` of the character that should be drawn for each cell in the matrix; this function essentially renders a simple bitmap matrix as unicode characters.

___

### **render.image()**

::: {.callout-tip title="code" collapse="true"}
```{r eval=F}
render.image = function(img,shades=2){
	#luma greyscale
	img = 0.2126 * img[,,1] + 
		0.7152 * img[,,2] + 
		0.0722 * img[,,3]
	
	#filter
	floor(img * shades) 
}
```
:::

Images are prepared for `render.matrix()` by flooring them to only a few integer values (determined by `shades`) to be used to index the character palette. Before this, the image is greyscaled by combining the color channels into one matrix.

___

### **load_video()**

::: {.callout-tip title="code" collapse="true"}
```{r eval=F}
load.video = function(path,size=128,framerate=25) {
	#size = desired horizontal resolution
	
	cat('Downscaling...\n')
	#get resolution metadata
	info = av_media_info(path)$video 
	
	scale = c(info$width,info$height)
	scale = scale * (size/scale[1]) #scale to size
	scale[2] = scale[2]/2 #halve vertical scale
	scale = floor(scale/2)*2 #make resolution even
		#the encoder hates odd resolutions
	
	#create temporary directory for downscaled video 
	temp = tempdir()
	temp = paste0(temp,'downsampled.mp4')
	av_encode_video(
		path,
		output = temp,
		framerate = framerate,
		vfilter = paste0("scale=",scale[1],':',scale[2])
	) #downscale video
	
	cat('Extracting frames...\n')
	av_video_images(temp,fps=framerate) 
		#automatically creates its own temporary directory
}
```
:::

Before using `av::av_video_images()` to generate a list of image files for `play.video()`, we use `av::av_encode_video()` to downscale the image to the desired resolution for rendering.

We could manually downsample the image files to get the desired resolution, but it's a much better idea to precompute this step, since generating the image files in the first place already requires precomputation.

The downscaled video is saved to a temporary file, and `av::av_video_images()` generates a temporary directory to store the image frames.

___

### **play_video()**

::: {.callout-tip title="code" collapse="true"}
```{r eval=F}
play.video = function(
		video_data,
		target_framerate = 25,
		palette=c('▓','∏','░',':',' ')
){
	frame = 1
	while (frame < length(video_data)){
		#get time before rendering frame
		t = as.double(Sys.time()) 
		
		#render frame
		render.matrix(
			render.image(
				readJPEG(video_data[frame]),
				shades=length(palette)
			),
			palette=palette
		)
		
		#how long that took to render
		dur = as.double(Sys.time()) - t 
		
		#how long that took in frames
		dur_frames = dur/(1/target_framerate) 
		
		#advance to next frame
		frame = frame + ceiling(dur_frames) 
		
		#time until the next frame
		buff = (dur_frames - floor(dur_frames)) * 
			1/target_framerate 
		
		#sleep to the start of the next frame
		Sys.sleep(1/target_framerate - buff) 
	}	
}
```
:::

`play.video` attempts to render each image every 1/`target_framerate` seconds.

The rendering is done with `render.matrix(render.image(readJPEG(file)))`, where `file` is the path to the image file supplied by `load_video()`.

In order to play at the desired framerate, the video player records the time it takes to read and render the frame (which is not insignificant, especially at higher resolutions) and then sleeps for the remainder of the frame.

The function also has code to skip frames if the time it took was greater than the intended duration of the frame.

## Notes

### Palette

I chose this default (recommended) palette after quite a bit of experimenting. It aims to provide high contrast while still giving good tone resolution without using too many shades; it's quite a bit more interpretable than many of the other palettes I tried.

The supply of good-quality characters for palettes is severely limited: many characters (notably, the full block) render too small or not monospaced[^1] in the RStudio console, and even those that don't still have significant gaps between lines. This is one of the few combinations with good contrast that doesn't introduce distracting shapes and patterns.

[^1]: Of the default uniform block shading elements █▓▒░, RStudio renders █ at half height and ▒ as thinner than a regular monospaced character. It's absurd!

---

### Resolution

`cat()` struggles to render quickly at higher resolutions (as well as on some devices in general). Reducing the resolution without losing a lot of visual clarity is challenging, especially with the limited palette choices. I experimented with generating characters with marching squares, but this proved to be far less effective than just varying the shades. This may be worth revisiting with cleverer use of the alphabet and combining both approaches.

------------------------------------------------------------------------

### Pure R

This is intended to run in RStudio; it actually does work in R.app, but R.app renders much less frequently than RStudio and therefore can't support a very good framerate.

---

### History

Rendering video in the console is a bit of a canonical exercise, and has been done many times before (though not in RStudio, as far as I can tell). The purpose of this bite is mainly to provide a very lightweight set of functions (~20 lines of code) to accomplish this.

