---
title: "Play Videos in R"
format: 
  html:
    toc: true
    toc-depth: 3
editor: visual
---

This bite has some quick code to parse video files and render them to the console using only vanilla methods. (and the `av` package to parse video)

```{r eval=F}
vid = load.video('XcQ.mp4')
play.video(vid)
```

![]() \# REMEMBER TO ADD GIF OF THIS IN ACTION

### Big Picture

To render a video, we need:

-   a way to convert high-res, color images to something viewable in the console

-   a way to load and display each frame at the right time

### Functions

------------------------------------------------------------------------

#### `render.matrix(matrix,palette)`

Renders an integer matrix to the console as a single line\*, overwriting the previous render.matrix() output. This allows us to animate video frames by drawing them in succession.

------------------------------------------------------------------------

#### `render.image(img,size,shades)`

Converts an image (e.g. from `jpeg::readJPEG()`) into a matrix suitable for `render.matrix`. This downsamples and quantizes it to only a few shades of grey.

------------------------------------------------------------------------

#### `load.video(path,size)`

Reads a video file and generates an image for each frame using `av::av_video_images()`. To expedite the rendering, the video is also downscaled.

------------------------------------------------------------------------

#### `play.video(video_data,size,palette)`

Reads a list of images as generated by `load.video()` or `av::av_video_images()` and renders each frame at the appropriate time. If rendering takes too long, frames will be skipped to preserve the overall video speed.

### Code Details

::: {.callout-tip title=`render.matrix` collapse="TRUE"}
```{r eval=F}
render.matrix = function(M,palette=c('▓',' ')) {
	cat('\r',           #clear previous render	
		c('\t',palette)[  #index palette to convert matrix to characters
			t(cbind(M,matrix(-1,nrow(M),10)))+2 
											#pad matrix with tab characters
		], sep='')
}
```
:::

Console rendering is achieved by calling `base::cat()` on a single-line string which wraps to match the dimensions of the matrix we feed in.

Additionally, we:

- use `\r` to clear the previous frame rendered by the function

- use `\t` tab characters to wrap each row onto a new line, *without* having to use actual `\n` newlines (which would break `\r`)

The input matrix consists of positive integers (and 0) corresponding to the index in `palette` of the character that should be drawn for each cell in the matrix; this function essentially renders a simple bitmap matrix as unicode characters.

___
::: {.callout-tip title=`render.matrix` collapse="TRUE"}
```{r eval=F}
render.image = function(img,size=128,shades=2){ #size = horizontal resolution
	by = ncol(img)/size #sample every this many pixels
	
	#nearest-neighbor downsample, with double horizontal resolution
	img = img[seq(1,nrow(img),by*2),seq(1,ncol(img),by),]
	
	#luma greyscale
	img = 0.2126 * img[,,1] + 
				0.7152 * img[,,2] + 
				0.0722 * img[,,3]
	
	#filter
	floor(img * shades)
}
```
:::


