

# INTRO NEEDED

and photos! of the R stuff especially, and a glam photo of the phone and mouse setup lol

START
BEGIN
HERE



**Disclaimer**: Back up your phone before doing this! I did this with an old flip phone that I factory reset, so I wasn't looking out for security or data concerns. Try at your own risk.


## 1. Termux: Android Linux Emulator

The first step is to install the [Termux](https://github.com/termux/termux-app) APK, which emulates a linux terminal with a decent amount of control.

::: {.callout-warning title="Termux Installation on my Flip Phone"}
I'm using an old flipphone that isn't really meant to run user-provided APKs. This was the process I figured out (from fragmentary advice online) to install and run arbitrary apks:

- enable dev mode by typing `*#*#33284#*#*` (i.e., `*#*#DEBUG#*#*` with the keypad)
- connect phone to computer via microusb, click OK on phone to enable usb file transfer
- use [https://app.webadb.com](https://app.webadb.com) to install the APK ("Install APK" on sidebar) via ADB interface. (you can also do the ADB process manually)

- (Note: APKs can be uninstalled by opening "Interactive Shell" from the sidebar, finding the app name by running `pm list packages`, and then running `pm uninstall -k --user 0 PACKAGENAME`)

Once I installed the Termux APPK this way, I had to set a main menu shortcut in the settings to be able to launch it.

(To make it easier to launch other APKs, I also installed [BlueLineConsole](https://github.com/nhirokinet/bluelineconsole) (other launchers didn't work on this device), which I quite like. For our R purposes we won't need any APKs other than Termux, though.)
:::



## 2. Running R in Termux[^guide]

[^guide]: I got most of this from [here](https://old.reddit.com/r/termux/comments/1ardg2l/installing_r_programming_language_issues/kqix0rg/). There's another resource you might see online for running R on Android, but it no longer works!

1. install and run the Termux app

2. ensure you have full keyboard control[^keyb] to use the terminal

[^keyb]: I had to connect a bluetooth keyboard since Termux didn't like the flip phone keypad. I believe the regular touchscreen keyboard (along with Termux's built-in buttons) should suffice for a normal android phone. 

```{r eval=F}
echo Hello World
```

3. install Debian in Termux:

```{r eval=F}
pkg install proot-distro
proot-distro install debian
```

4. log into Debian and install R and some useful commands[^comms]

[^comms]: Only `r-base-dev` is required to run R, but the other two are useful to have around.

```{r eval=F}
proot-distro login debian 
apt install curl
apt install git
apt install r-base-dev
```

5. you can now run a regular old R session!

```{r eval=F}
R
print(R.version.string)
```

6. install.packages() generally works for packages that don't build their own stuff on install.

```{r eval=F}
install.packages("dplyr")
library(dplyr)
glimpse(iris,32)
```

:::{.callout-note title="Package Compatibility" collapse="true"}
Tested and working:

- `dplyr`

Failed to install:

- `devtools`
- `httr2` (you can get around this with clever use of `system()` and curl)
- `httpuv`
:::

